D3064R0: OOTA Execution on Real Implementations is Provably Vacuous

### 2023-12-18 ootavacuous.tex
Start appendices on new page and transcribe todo lists
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Update x86 coe example to more typical scenario
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Add diagrams explaining temporal properties of coe, fre, and rfe
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Propagation delay and control dependencies
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

Here there is a semantic dependency from \co{x}, but in some executions
it extends to \co{y} and in others to \co{z}.

@@

\appendix

@@

This appendix evaluates a number of OOTA-related litmus tests.
The evaluations are carried out using the \co{herd7} tool and also
using manual analysis.
The manual analysis uses the informal rule that a semantic dependency
exists between a given relaxed atomic load and store on any execution
where a different value loaded would change either the address or the
value of that store.
Producing a precise formal statement of this rule is quite challenging.

@@

\begin{figure}[tb]
\begin{center}
\resizebox{5in}{!}{\includegraphics{coe-out-4}}
\caption{On x86, coe Links Are Partially Ordered}
\label{fig:On x86, coe Links Are Partially Ordered}
\end{center}
\end{figure}

@@

And this is exactly why the C++ memory model guarantees ordering from
rfe links but not the coe and fre links for relaxed accesses in the
absence of other ordering from stronger atomic memory accesses or
\co{atomic_thread_fence()}.

@@

% @@@ TODO:
% Measure store-buffer delays on an x86 system.
% Real-world compilers freeze nondeterministic choices.

>>>>

Here there is a semantic dependency from \co{x}, but in some executions
it extends to \co{y} and in others to \co{z}.
This is an example of the load affecting not the value or address of
a given store, but rather whether or not that store is executed at all.

@@

\clearpage
\appendix

@@

This appendix evaluates a number of OOTA-related litmus tests.
The evaluations are carried out using the \co{herd7} tool and also
using manual analysis.
The manual analysis uses the informal rule that a semantic dependency
exists between a given relaxed atomic load and store on any execution
where a different value loaded would change the address stored to, the
value stored, or the whether or not that store would be executed in
the first place.
Producing a precise formal statement of this rule turns out to be quite
challenging.

@@

\begin{figure}[tb]
\begin{center}
\resizebox{5in}{!}{\includegraphics{coe-out-1}}
\caption{On x86, coe Links Are Partially Ordered}
\label{fig:On x86, coe Links Are Partially Ordered}
\end{center}
\end{figure}

@@

\begin{figure}[tb]
\begin{center}
\resizebox{4in}{!}{\includegraphics{coe}}
\resizebox{4in}{!}{\includegraphics{fre}}
\resizebox{4in}{!}{\includegraphics{rfe}}
\caption{Propagation Delay and Temporal Properties of coe, fre, and rfe}
\label{fig:Propagation Delay and Temporal Properties of coe, fre, and rfe}
\end{center}
\end{figure}

Figure~\ref{fig:Propagation Delay and Temporal Properties of coe, fre, and rfe}
shows how the temporal properties of coe, fre, and rfe are explained
by propagation delay.
In the case of coe, the two stores take time to propagate across the
system, and their ordering can and is decided after the fact.
In the case of fre, the store again takes time to propagate across
the system, and so a load might execute later in global time than
did the store, but not late enough to return that store's value.
In the case of rfe, the load cannot return a given store's value until
the value has propagated to that part of the system executing that load.

And this is exactly why the C++ memory model guarantees ordering from
rfe links but not the coe and fre links for relaxed accesses in the
absence of other ordering from stronger atomic memory accesses or
\co{atomic_thread_fence(memory_order_seq_cst}.

@@

% @@@ TODO:
% Convert remaining PPCMEM diagrams to herd7
% Emphasize that real-world implementations map abstract machine to matter
%	and energy in the real world.
% Look into why inkscape no longer likes generating PDFs.
